Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2021-11-18 hpp.tex
Final version before submission
  Remove color blue for text that had been added,
  Add bios.

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{{\color{blue} Example of manipulation planning problem.} Top: two UR3 robots with one gripper each (X=red, Y=green, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for example, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}

@@

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that can handle implicit and explicit constraints,
\item a manipulation planning algorithm that tackles a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above, following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found at \href{https://humanoid-path-planner.github.io/hpp-doc}{https://humanoid-path-planner.github.io/hpp-doc}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only mentioned in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results for several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces nonlinear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \Partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

@@

Some systems are subject to nonlinear constraints. These
constraints define submanifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping the same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another method consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this can be done. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference lies in the fact that the nodes built on the
tangent space are not immediately projected onto the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} put forward a general framework to plan task-constrained
motions in the presence of moving obstacles. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

The first attempt to solve manipulation planning problems using random sampling
was proposed by {\color{blue}Siméon \textit{et al}}~\cite{simeon2004ijrr} where a reduction property simplifies the problem.

@@

{\color{blue}Schmitt \textit{et al}}~\cite{schmitt19icra} propose an approach where two robots manipulate an object in a dynamic environment. The output of the algorithm is a sequence of controllers rather than a sequence of paths.

Our work shares many ideas with {\color{blue}Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to the latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We only address prehensile manipulation.

@@

Open-source software platforms are an important tool to enable fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. Undoubtedly the most popular one is OMPL \cite{ompl2012} which integrates many randomized path planning algorithms and is widely used for teaching purposes. Recently, {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to nonlinear constraints.

@@

The joint configuration spaces listed in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{\color{blue}Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. Here we detail only those properties that are useful for the following developments.

@@

Elements of $SE(2)$ are represented by a vector of dimension 4. The first 2 components represent the image of the origin. The last 2 components represent the cosine and sine of the rotation angle. Therefore the homogeneous matrix associated to $\conf=(q_1,q_2,q_3,q_4)$ is
$$
M = \left(\begin{array}{lll}
  q_3 & -q_4 & q_1 \\
  q_4 &  q_3 & q_2 \\
  0   &  0  & 1
\end{array}\right).
$$
Table~\ref{tab:vector representation} compiles this information.
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
    Lie group type & configuration & velocity\\
    \hline
    $SE(3)$ & $(x_1,x_2,x_3,p_1,\cdots,p_4)$ & $\dconf=(\mathbf{v},\omega)$ \\
            & $\in\reals^7$                 & $\in\reals^6$ \\
    $SE(2)$ & $(x_1,x_2,\cos\theta,\sin\theta)$ & $\dconf=(\mathbf{v},\dot{\theta})$\\
            & $\in\reals^4$                 & $\in\reals^3$ \\
    $SO(3)$ & $(p_1,p_2,p_3,p_4)\in\reals^4$ & $\dconf=\omega\in\reals^3$\\
    $SO(2)$ & $(\cos\theta,\sin\theta)\in\reals^2$ & $\dconf=\dot{\theta}\in\reals$\\
    \hline
  \end{tabular}
  \caption{Main Lie group types and their vector representations. Notice that the dimensions of the configuration representation and of the velocity representation may differ. {\color{blue} Using $(\cos\theta,\sin\theta)$ instead of $\theta$ for $SO(2)$ and $SE(2)$ makes the parameterization continuous when $\theta$ discontinuously switches from $-\pi$ to $\pi$}}
  \label{tab:vector representation}
\end{table}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the Cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the {\color{blue}starting} indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 {\color{blue}\ominus} \conf_0)
$$
This formula generalizes the linear interpolation to robots with free-flying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

\begin{definition}\textbf{nonlinear constraint.} A nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non-zero right hand side for the same function $h$. For that we define parameterized nonlinear constraints.
\begin{definition}\textbf{Parameterized nonlinear constraint.}
  A parameterized nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{\color{blue} Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a {\color{blue} piece-wise} differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the submanifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\color{blue}\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\color{blue}\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{who has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ solves the linear approximation, but it may not be the best choice in general.

The computation of $\alpha_i$ is performed by a line search algorithm.
The algorithm stops when the norm of each $h_i(\conf_{i+1})$ is below a given error threshold.
%% begin impl
Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1HierarchicalIterative.html}{\texttt{HierarchicalIterative}} implements the above Newton method.
{\color{blue}
Several line search methods are implemented:
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1Backtracking.html}{\texttt{Backtracking}} \cite{backtracking},
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1ErrorNormBased.html}{\texttt{ErrorNormBased}}:
  $$\alpha_{i} = C - K  \text{tanh}(a \frac{\|f(\mathbf{q}_i)\|}{\epsilon^2} + b),$$
  where $C$, $K$, $a$, and $b$ are constant values, and $\epsilon$ is the
  error threshold,
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1FixedSequence.html}{\texttt{FixedSequence}} implements a fixed sequence of $\alpha_i$ that converges to 1,
\item and \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1Constant.html}{\texttt{Constant}} sets $\alpha_i$ to 1.
\end{itemize}
}
Note that to define a new constraint, the user needs to derive class \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}} and to implement methods \texttt{impl\_compute} and \texttt{impl\_jacobian}.
%% end impl

@@

Instead, we propose an alternative architecture where paths store the constraints they are subject to and apply the constraints at path evaluation {\color{blue}(\textit{i.e.} when computing the configuration at a given parameter)}. Let $P\in C^1([0,T],\CS)$ be a path without constraint defined on an interval $[0,T]$, and $\mathbf{proj}$ a projector onto a submanifold defined by numerical constraints (\textit{i.e.} an instance of\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}).

@@

Projecting configurations at path evaluation has the advantage of not losing information. In return, the projection of a continuous path may be discontinuous. Thus, before inserting a projected path in a roadmap for example, it is necessary to detect possible discontinuities. {\color{blue}Hauser} \cite{Hauser-RSS-13} proposes a solution to this problem. {\color{blue}In a previous paper} \cite{mirabel:hal-01360409}, {\color{blue}we} described two algorithms to check whether a projected path is continuous. These algorithms are implemented by
classes \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1pathProjector_1_1Dichotomy.html}{\texttt{pathProjector::Dichotomy}} and \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1pathProjector_1_1Progressive.html}{\texttt{pathProjector::Progressive}}. {\color{blue} Note that when a path is not continuous, the algorithms return a continuous portion of the path starting at the beginning of the path. This enables function \EXTEND in Algorithm~\ref{algo:M-RRT} to create a new node.}

@@

    \If {$p$ collision-free \color{blue}and continuous} {\Return $p$}
    \Else { \Return collision-free {\color{blue}continuous} portion of $p$ starting at $\conf_{near}$}

>>>>

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{{ Example of manipulation planning problem.} Top: two UR3 robots with one gripper each (X=red, Y=green, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for example, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}

@@

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots. Figure~\ref{fig:2ur3-cylinder} displays an example of manipulation problem.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that can handle implicit and explicit constraints,
\item a manipulation planning algorithm that tackles a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above, following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{ Installation instructions can be found at \href{https://humanoid-path-planner.github.io/hpp-doc}{https://humanoid-path-planner.github.io/hpp-doc}.}
This paper extends the work presented in { previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only mentioned in {Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results for several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces nonlinear constraints and solvers that are at the core of the manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \Partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

@@

Some systems are subject to nonlinear constraints. These
constraints define submanifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping the same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another method consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this can be done. {Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference lies in the fact that the nodes built on the
tangent space are not immediately projected onto the manifold.
{Cefalo \textit{et al}}~\cite{CefOri2019} put forward a general framework to plan task-constrained
motions in the presence of moving obstacles. {Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

The first attempt to solve manipulation planning problems using random sampling
was proposed by {Siméon \textit{et al}}~\cite{simeon2004ijrr} where a reduction property simplifies the problem.

@@

{Schmitt \textit{et al}}~\cite{schmitt19icra} propose an approach where two robots manipulate an object in a dynamic environment. The output of the algorithm is a sequence of controllers rather than a sequence of paths.

Our work shares many ideas with {Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to the latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We only address prehensile manipulation.

@@

Open-source software platforms are an important tool to enable fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. Undoubtedly the most popular one is OMPL \cite{ompl2012} which integrates many randomized path planning algorithms and is widely used for teaching purposes. Recently, {Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to nonlinear constraints.

@@

The joint configuration spaces listed in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. Here we detail only those properties that are useful for the following developments.

@@

Elements of $SE(2)$ are represented by a vector of dimension 4. The first 2 components represent the image of the origin. The last 2 components represent the cosine and sine of the rotation angle. Therefore the homogeneous matrix associated to $\conf=(q_1,q_2,q_3,q_4)$ is
$$
M = \left(\begin{array}{lll}
  q_3 & -q_4 & q_1 \\
  q_4 &  q_3 & q_2 \\
  0   &  0  & 1
\end{array}\right).
$$
Table~\ref{tab:vector representation} compiles this information.
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
    Lie group type & configuration & velocity\\
    \hline
    $SE(3)$ & $(x_1,x_2,x_3,p_1,\cdots,p_4)$ & $\dconf=(\mathbf{v},\omega)$ \\
            & $\in\reals^7$                 & $\in\reals^6$ \\
    $SE(2)$ & $(x_1,x_2,\cos\theta,\sin\theta)$ & $\dconf=(\mathbf{v},\dot{\theta})$\\
            & $\in\reals^4$                 & $\in\reals^3$ \\
    $SO(3)$ & $(p_1,p_2,p_3,p_4)\in\reals^4$ & $\dconf=\omega\in\reals^3$\\
    $SO(2)$ & $(\cos\theta,\sin\theta)\in\reals^2$ & $\dconf=\dot{\theta}\in\reals$\\
    \hline
  \end{tabular}
  \caption{Main Lie group types and their vector representations. Notice that the dimensions of the configuration representation and of the velocity representation may differ. { Using $(\cos\theta,\sin\theta)$ instead of $\theta$ for $SO(2)$ and $SE(2)$ makes the parameterization continuous when $\theta$ discontinuously switches from $-\pi$ to $\pi$}}
  \label{tab:vector representation}
\end{table}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the Cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the {starting} indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 {\ominus} \conf_0)
$$
This formula generalizes the linear interpolation to robots with free-flying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

\begin{definition}\textbf{nonlinear constraint.} A nonlinear constraint is defined by a { piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non-zero right hand side for the same function $h$. For that we define parameterized nonlinear constraints.
\begin{definition}\textbf{Parameterized nonlinear constraint.}
  A parameterized nonlinear constraint is defined by a { piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{ Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a { piece-wise} differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the submanifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{who has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ solves the linear approximation, but it may not be the best choice in general.

The computation of $\alpha_i$ is performed by a line search algorithm.
The algorithm stops when the norm of each $h_i(\conf_{i+1})$ is below a given error threshold.
%% begin impl
Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1HierarchicalIterative.html}{\texttt{HierarchicalIterative}} implements the above Newton method.
{
Several line search methods are implemented:
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1Backtracking.html}{\texttt{Backtracking}} \cite{backtracking},
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1ErrorNormBased.html}{\texttt{ErrorNormBased}}:
  $$\alpha_{i} = C - K  \text{tanh}(a \frac{\|f(\mathbf{q}_i)\|}{\epsilon^2} + b),$$
  where $C$, $K$, $a$, and $b$ are constant values, and $\epsilon$ is the
  error threshold,
\item \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1FixedSequence.html}{\texttt{FixedSequence}} implements a fixed sequence of $\alpha_i$ that converges to 1,
\item and \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/structhpp_1_1constraints_1_1solver_1_1lineSearch_1_1Constant.html}{\texttt{Constant}} sets $\alpha_i$ to 1.
\end{itemize}
}
Note that to define a new constraint, the user needs to derive class \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}} and to implement methods \texttt{impl\_compute} and \texttt{impl\_jacobian}.
%% end impl

@@

Instead, we propose an alternative architecture where paths store the constraints they are subject to and apply the constraints at path evaluation {(\textit{i.e.} when computing the configuration at a given parameter)}. Let $P\in C^1([0,T],\CS)$ be a path without constraint defined on an interval $[0,T]$, and $\mathbf{proj}$ a projector onto a submanifold defined by numerical constraints (\textit{i.e.} an instance of\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}).

@@

Projecting configurations at path evaluation has the advantage of not losing information. In return, the projection of a continuous path may be discontinuous. Thus, before inserting a projected path in a roadmap for example, it is necessary to detect possible discontinuities. {Hauser} \cite{Hauser-RSS-13} proposes a solution to this problem. {In a previous paper} \cite{mirabel:hal-01360409}, {we} described two algorithms to check whether a projected path is continuous. These algorithms are implemented by
classes \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1pathProjector_1_1Dichotomy.html}{\texttt{pathProjector::Dichotomy}} and \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1pathProjector_1_1Progressive.html}{\texttt{pathProjector::Progressive}}. { Note that when a path is not continuous, the algorithms return a continuous portion of the path starting at the beginning of the path. This enables function \EXTEND in Algorithm~\ref{algo:M-RRT} to create a new node.}

@@

    \If {$p$ collision-free and continuous} {\Return $p$}
    \Else { \Return collision-free {continuous} portion of $p$ starting at $\conf_{near}$}
